Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-05-06 manu-draft.qmd
added new methods text for gams, object for secchi on bottom

load(file = here('data/speidat.RData'))
load(file = here('data/gamfit.RData'))
load(file = here('tabs/dattab.RData'))
load(file = here('tabs/temptrndtab.RData'))
load(file = here('tabs/saltrndtab.RData'))
load(file = here('tabs/mixdaytab.RData'))
load(file = here('tabs/modtab.RData'))
load(file = here('data/sgmods.RData'))
```

@@

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years. Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely stressing seagrasses beyond their optimal tolerance ranges, particularly in upper bay segments where marginal light environments exist. This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community. While other studies have suggested negative effects from climate change in the form of increased salinity [e.g., @costa2023], lower salinity may be a stressor for seagrass in Tampa Bay. The analysis was supported by 30-year seagrass datasets including aerial surveys of total seagrass distribution, annual transect monitoring describing species-specific percent cover, and synoptic seagrass data collected with routine biotic and water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.

@@

Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a). Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014]. Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small, channelized tributaries and manmade flood control conveyances [@tbep0623]. Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay, causing relatively longer residence times and accumulation of legacy pollutant loads compared to the other bay segments [@sherwood2015; @luther2022]. Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023]. By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623]. All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m. Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments [@greening2014; @johansson15].

@@

The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017]. From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage. Though Tampa Bay was far from pristine in the 1950s, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage. Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a). External nitrogen load estimates decreased by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].

From 2016 to present, dramatic seagrass loss has been observed in Tampa Bay, despite nutrient loading and light attenuation remaining relatively stable (Figure [-@fig-seagrasschg]a). Total cover in Tampa Bay has decreased by 28% (11,518 acres/4,661 ha) from the 2016 peak to a total baywide coverage of 30,137 acres (12,196 ha) in 2022. Losses have been most pronounced in Old Tampa Bay (62%; 6,963 acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599 acres/647 ha loss). The current estimate for Old Tampa Bay of 4,183 acres (1,693 ha) is the lowest ever recorded in that bay segment since mapping efforts began in the 1980s. Coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss), whereas coverage in Lower Tampa Bay has remained stable, with only a 2% loss which is close to the mapping error. As such, trajectories of recovery and decline have varied by bay segment in magnitude and timing of the change, although a consistent decline baywide has been observed since 2016. Change by species has been most notable for *Halodule wrightii* (shoal grass) in OTB and HB, whereas some losses have been observed for *Thalassia testudinum* (turtle grass) in LTB, likely related to dominance of each species across the salinity gradient and proximity to hydrologic inputs [@lewis1985].

@@

The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity. The conceptual model for evaluating these changes describes the fundamental niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972]. In the simplest form, this can be conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter, where reduced growth or mortality is observed outside of these ranges. Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]). Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range. Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept). Although the fundamental niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been relatively stable in recent years [@tbep0323].

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the annual summaries. As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre 2016, decline post 2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).

Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables. Interactions with time period and a main effect for bay segment were also included as above. GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. The seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrasses. Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.

@@

```{r}
epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
fimmodsum <- modtxt_fun(sgmods$fimmod)
```

Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was influenced by climate stressors. The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (`r epcmodsum1`). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (`r epcmodsum2`), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.

The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to larger spread of the data (`r fimmodsum`). Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature or salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab). The interaction between salinity and temperature was not included as a significant term in the final FIM model. Lastly, the most parsimonious model for the PDEM dataset only included time period as a marginally significant variable and the overall model fit was not significant (@tbl-modtab).

@@

Additional limitations of our models may relate to an incomplete description of factors influencing seagrass growth, such as the inclusion of additional drivers and an incomplete or overly simplified causal network. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of external nitrogen inputs in affecting light environments for seagrass growth [@greening2014; @sherwood2017]. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity. However, our use of time period (pre/post 2016) as a categorical variable indirectly addressed this issue. Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass, i.e., modeling challenges related to correlated predictors [@fourqurean2003]. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay, particularly in OTB where the most loss occurred [@tbep0696; @Beck20a]. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be lower in locations where seagrasses are present and can absorb solar radiation, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses [@carlson2018]. This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors are a more likely scenario. This is especially true for water temperature trends captured by the EPC dataset which includes deeper, fixed sites adjacent to shallow seagrass flats.

@@

![Seagrass changes over time in Tampa Bay for (a) areal coverage 1988 - 2022 from mapping, (b) frequency occurrence of major species 1998 - 2022 from annual transect monitoring, and (c) mean annual light attenuation from 1970 to 2022. Changes are shown for major bay segments. Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres and red lines in (c) show the approximate linear trend for the period of record covered by the transect data in (b). The dashed line in (c) shows the light attenuation threshold for each bay segment identified as supportive for seagrass growth. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}

>>>>

load(file = here('data/speidat.RData'))
load(file = here('data/gamfit.RData'))
load(file = here('tabs/dattab.RData'))
load(file = here('tabs/temptrndtab.RData'))
load(file = here('tabs/saltrndtab.RData'))
load(file = here('tabs/mixdaytab.RData'))
load(file = here('tabs/modtab.RData'))
load(file = here('data/sgmods.RData'))
load(file = here('data/onb.RData'))
```

@@

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years. Three datasets of varying sampling designs and temporal coverage were used to assess the primary hypothesis that Tampa Bay is trending towards hotter and fresher conditions that are likely stressing seagrasses beyond their optimal tolerance ranges, particularly in upper bay segments that are more shallow and receive a majority of hydrologic inflow. This hypothesis was generated from preliminary assessments of datasets used herein and discussions with the regional scientific and management community. While other studies have suggested negative effects from climate change in the form of increased salinity [e.g., @costa2023], lower salinity may be a stressor for seagrass in Tampa Bay. The analysis was supported by 30-year seagrass datasets including aerial surveys of total seagrass distribution, annual transect monitoring describing species-specific percent cover, and synoptic seagrass data collected with routine biotic and water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable.

@@

Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a). Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014]. Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia Rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small, channelized tributaries and manmade flood control conveyances [@tbep0623]. Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay, causing relatively longer residence times and accumulation of legacy pollutant loads compared to the other bay segments [@sherwood2015; @luther2022]. Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023]. By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623]. All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m. Light penetration typically reaches bottom habitats under current conditions (Figure [-@fig-seagrasschg]c), although seagrasses were historically limited by high phytoplankton production that affected light environments [@greening2014; @johansson15].

@@

The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017]. From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage. Though Tampa Bay was far from pristine in the 1950s, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage. Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a). Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed \[@greening2014; @beck2019\].

From 2016 to present, dramatic seagrass loss has been observed in Tampa Bay, despite light environments remaining supportive of growth (Figure [-@fig-seagrasschg]c). Total cover in Tampa Bay has decreased by 28% (11,518 acres/4,661 ha) from the 2016 peak to a total baywide coverage of 30,137 acres (12,196 ha) in 2022. Losses have been most pronounced in Old Tampa Bay (62%; 6,963 acres/2,818 ha loss) and Hillsborough Bay (80%; 1,599 acres/647 ha loss). The current estimate for Old Tampa Bay of 4,183 acres (1,693 ha) is the lowest ever recorded in that bay segment since mapping efforts began in the 1980s. Coverage in Middle Tampa Bay decreased by 20% (1,926 acres/779 ha loss), whereas coverage in Lower Tampa Bay has remained stable, with only a 2% loss which is close to the mapping error. As such, trajectories of recovery and decline have varied by bay segment in magnitude and timing of the change, although a consistent decline baywide has been observed since 2016. Change by species has been most notable for *Halodule wrightii* (shoal grass) in OTB and HB, whereas some losses have been observed for *Thalassia testudinum* (turtle grass) in LTB, likely related to dominance of each species across the salinity gradient and proximity to hydrologic inputs [@lewis1985].

@@

The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity. The conceptual model for evaluating these changes describes the fundamental niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972]. In the simplest form, this can be conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter, where reduced growth or mortality is observed outside of these ranges. Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]). Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range. Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept). Although the fundamental niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been sufficient for growth in recent years (Figure [-@fig-seagrasschg]c).

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the annual summaries. As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment. GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both. Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696]. For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom. A single smooth term was used for each predictor using a thing plate regression spline, including a tensor product smoother term that evaluated the potential interacting effects of each predictor with year. Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).

Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GAMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables. Interactions with year for each predictor variable and separate smoothed terms for each bay segment were also included as above. GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses. Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects. Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models. For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.

Add equations.....

@@

```{r}
# epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
# epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
# fimmodsum <- modtxt_fun(sgmods$fimmod)
```

Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was influenced by climate stressors. The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.

The most parsimonious model for the FIM dataset also showed potential temperature and salinity associations with seagrass change after 2016, although explanatory power was lower than the EPC models likely due to larger spread of the data (). Increases in temperature and decreases in salinity were both associated with reductions in percent cover after 2016, whereas no relationships were observed prior to 2016 (p \< 0.05 and p \< 0.1 for the interactions between temperature or salinity with time period, respectively, Figure [-@fig-sgmod]b, @tbl-modtab). The interaction between salinity and temperature was not included as a significant term in the final FIM model. Lastly, the most parsimonious model for the PDEM dataset only included time period as a marginally significant variable and the overall model fit was not significant (@tbl-modtab).

@@

Additional limitations of our models may relate to an incomplete description of factors influencing seagrass growth, such as the inclusion of additional drivers and an incomplete or overly simplified causal network. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of external nitrogen inputs in affecting light environments for seagrass growth [@greening2014; @sherwood2017]. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity. However, our use of time period (pre/post 2016) as a categorical variable indirectly addressed this issue. Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass, i.e., modeling challenges related to correlated predictors [@fourqurean2003]. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has improved over the period of record and within the limits estimated to be supportive of seagrass growth in Tampa Bay [@tbep0400], particularly in OTB where the most loss occurred (Figure -[@fig-seagrasschg]c). A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be lower in locations where seagrasses are present and can absorb solar radiation, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses [@carlson2018]. This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors are a more likely scenario. This is especially true for water temperature trends captured by the EPC dataset which includes deeper, fixed sites adjacent to shallow seagrass flats.

@@

![Seagrass changes over time in Tampa Bay for (a) areal coverage 1988 - 2022 from mapping, (b) frequency occurrence of major species 1998 - 2022 from annual transect monitoring, and (c) mean annual light attenuation from 1970 to 2022. Changes are shown for major bay segments. Red lines in (a) show approximate capacity of seagrass coverage based on the baywide target of 40,000 acres and red lines in (c) show the approximate linear trend for the period of record covered by the transect data in (b). The dashed line in (c) shows the light attenuation threshold to support seagrass growth in each bay segment. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay, LTB: Lower Tampa Bay.](figs/seagrasschg.png){#fig-seagrasschg}
